---
title: "Illustration of Functional Forms in Joint Models using JMbayes2"
author: "Companion code for: Functional forms in joint models for longitudinal and time-to-event data"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
---

# Overview

This document illustrates the syntax for fitting the different functional forms
described in the accompanying paper

> **Functional forms in joint models for longitudinal and time-to-event data: A practical guide with application and interpretation**

using the example **Primary Biliary Cirrhosis (PBC)** dataset included in the
`JMbayes2` package.

The purpose of this document is to demonstrate the specification of the
different association structures. The analyses presented here are illustrative
only and are **not** intended to reproduce the results of the paper, which were
based on the MIRAGE glioblastoma study.

```{r setup, message=FALSE, warning=FALSE}
library(JMbayes2)
library(nlme)
library(survival)
library(splines)

data("pbc2")
data("pbc2.id")

pbc2.id$status2 <- as.numeric(pbc2.id$status != "alive")
```

# Step 1: Fit the longitudinal submodel

We model the longitudinal evolution of log serum bilirubin using a linear
mixed-effects model with a natural cubic spline for time.

```{r}
fit_mixed <- lme(
  log(serBilir) ~ ns(year, 2),
  random = ~ ns(year, 2) | id,
  data = pbc2
)
```

# Step 2: Fit the survival submodel

A Cox proportional hazards model is used for the event process.

```{r}
fit_cox <- coxph(
  Surv(years, status2) ~ sex,
  data = pbc2.id,
  x = TRUE
)
```

# Step 3: Fit the joint models

## Current value

```{r eval=FALSE}
jm_current <- jm(
  fit_cox,
  fit_mixed,
  time_var = "year",
  n_chains = 5,
  n_iter = 5000,
  n_burnin = 1000
)

summary(jm_current)
```

---

## Time-dependent slope

```{r eval=FALSE}
jm_slope <- update(
  jm_current,
  functional_forms = ~ velocity(log(serBilir))
)

summary(jm_slope)
```

---

## Acceleration

```{r eval=FALSE}
jm_acceleration <- update(
  jm_current,
  functional_forms = ~ acceleration(log(serBilir))
)

summary(jm_acceleration)
```

---

## Cumulative effect

```{r eval=FALSE}
jm_auc <- update(
  jm_current,
  functional_forms = ~ area(log(serBilir))
)

summary(jm_auc)
```

---

## Restricted cumulative effect

The hazard depends on the average biomarker exposure over the previous year.

```{r eval=FALSE}
jm_auc_recent <- update(
  jm_current,
  functional_forms =
    ~ area(log(serBilir), time_window = 1)
)

summary(jm_auc_recent)
```

---

## Change from baseline

```{r eval=FALSE}
jm_delta <- update(
  jm_current,
  functional_forms = ~ Delta(log(serBilir))
)

summary(jm_delta)
```

---

## Change over the previous year

```{r eval=FALSE}
jm_delta_recent <- update(
  jm_current,
  functional_forms =
    ~ Delta(log(serBilir), time_window = 1)
)

summary(jm_delta_recent)
```

---

<h2>References</h2>

Rizopoulos D, Miranda Afonso P, Papageorgiou G. **JMbayes2: Extended Joint Models for Longitudinal and Time-to-Event Data**. R package version 0.5-95; 2025. Available at: <https://github.com/drizopoulos/jmbayes2>.